Optimal analysis of Discrete-time Time-Invariant Affine Systems

### 2018-10-16 paper.tex
derniere passe

We can represent the evolution of an affine system by the following relation:
\begin{equation}
\label{context}
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k + b  
\end{equation}
where:
\begin{itemize}
\item $A$ is a non-zero square matrix of size $d\times d$;
\item $b$ is vector of $\rd$;
\item  $\xin$ is a non-empty polytope (bounded polyhedra). 
\end{itemize}
We insist on the fact that the initial values are represented by an infinite bounded set. It can be interpreted as a set of perturbations, non-determinism, different cases...

@@

For the matrix $A$ of the affine system considered at Eq.~\eqref{context}, we introduce the set $\lyap{A}$ of the  solutions  of the discrete Lyapunov equation :
\[
\lyap{A}=\{P\in\sym \mid P\succ 0,\ P-A^\intercal P A\succ 0\}\enspace .
\]
Hence, $\rho(A)<1$ ensures that $\lyap{A}$ is non-empty. More precisely, $\rho(A)<1$ if and only if for all positive definite matrices $R$ there exists a unique positive definite matrix $P$ such that $P-A^\intercal P A=R$. \ForAuthors{reference sur stein equation}

@@

We recall Weyl's inequalities~\cite{horn1990matrix} that provide inequalities for the eigenvalues of the sum of two symmetric matrices and the sum of the eigenvalues of each matrix.  
\begin{lemma}[Weyl's inequalities]
\label{le}
Let $M$ and $N$ be symmetric matrices. We have, for all $\ell\in\{1,\ldots,d\}$ :
\[
\lambda_\ell(M)+\lmin{N}\leq \lambda_\ell(M+N)\leq \lambda_\ell(M)+\lmax{N}
\]
\end{lemma}
The following lemma recalls that, when we maximize a convex function over a polytope, it suffices to consider the values of the function at the extreme points. 
\begin{lemma}[Maximisation of a convex function over a polytope]
\label{lemma2}
Let $C$ be a polytope and $f:\rd\to\rd$ a convex function. Then:
\[
\max_{x\in C} f(x)=\max_{x\in \mathcal{E}(C)} f(x)
\]
where $\mathcal{E}(C)$ denotes the finite set of the extreme points (vertices) of $C$.
\end{lemma}
Note that the result also holds for convex compacts sets.  
\subsection{Verification Linear Time-Invariant Discrete-time Systems}
\label{mainsub}
Now we come back to the verification problem presented at~\eqref{pbopt}. In this subsection, we suppose that the system in linear i.e. $b=0$ in~\eqref{context}. We will show at Subsection~\ref{affine} that the case $b\neq 0$ can be reduced to the linear case.

@@

\ForAuthors{Compare with zero in the interior of xin... if 0 belongs to the interior of xin then we can choose xeps=epsilon sign(q) in xin and then q.xeps >0 and xeps Q xeps nonnegative...}

\ForAuthors{Q cannot be null indeed for all k the value would be negative or null}

\ForAuthors{From double linear inequalities i.e. alpha<=cx<= beta, we can cast into a quadratic inequality. However if the scalar part is not negative alpha*beta, we cannot use our method. }  
We will need the following notation :
\[
\ks:=\inf\left\{k\in\nn\mid \min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\right\}>0\right\}
\]
and
\[
\lfrak:= \min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^{\ks})^\intercal Q A^{\ks}k x+q^\intercal A^{\ks}x\right\}
\]

@@

@@

@@

@@

%&=&\displaystyle{\max\left\{\max_{0\leq k\leq \bigk-1}\max_{x\in \mathcal{E}(\xin)} \nu_k(x),\lfrak\right\}}
%\end{array}
%\]
Now, for all $k< \ks$, for all $x\in\xin$, $\nu_k(x)\leq 0$ and for all $x\in\xin$, $\nu_{\ks}(x)\leq \lfrak$. Moreover, from Th~\ref{thfond}, 
for all $k>k^*$, $\nu_k(x)\leq \lfrak$. The conclusion follows from $\displaystyle{\max_{x\in\xin} \max_{\ks\leq k\leq k^*} \nu_k(x)\geq \max_{x\in\xin} \nu_{\ks}(x)\geq \lfrak}$. 
%The conclusion follows from the fact that  $\lfrak\leq \max_{x\in \mathcal{E}(\xin)} \nu_{\ks}(x)$.
\end{proof}
The integer $\bigk$ is the best integer than we can get in theory. In practice, we cannot compute this integer since we cannot solve the minimization problems :
\[
\inf_{P\in\lyaun} K(1,P)\text{ and }\inf_{P\in\lyap{A}}K(\lmax{P^{-1/2} Q P^{-1/2}},P)
\]
In practice, for the moment, we only compute two integers : $K( 1,P)$ where $P$ is an optimal solution of a perturbation of the minimization problem : 
\begin{equation}
\label{minlmaxQ}
\inf\{\lmax{P}\mid P\in \lyaun\}
\end{equation}
and $K(\lmax{P^{-1/2} Q P^{-1/2}},P)$ where $P$ is an optimal solution of a perturbation of the minimization problem :
\begin{equation}
\label{minlyap}
\inf\{\lmax{P}\mid P\in\lyap{A}\}\enspace .
\end{equation}
Then we take the minimum of the  two integers. We come back latter on the perturbation of the two minimization problems at Section~\ref{computations}.
\subsection{From linear systems to affine ones}
\label{affine}
We come back to Problem~\eqref{context}: the case where the system is affine ($b\neq 0$).
A possible natural reformulation of affine systems into linear one is to use lift-and-project. However, lift-and-project would add 1 as eigenvector of the matrix defining the new linear system. Assumption~\ref{assum2} would be violated.  Consequently, we adopt another basic approach which consists in using an auxiliary linear discrete-time system.

 We recall that we have :
%Let us consider the same problem where we replace the linear dynamics by an affine one:
\[
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k+b,\
\]
where $A$ is a $d\times d$ matrix and $b$ a vector of $\rd$. We recall that $\rea$ is given by Formula~\eqref{reach} and we consider the optimization problem:
\[
\sup_{x\in\rea} x^\intercal Q x+q^\intercal x
\]
Assumption~\ref{assum2} still holds and it implies that $Id-A$ is invertible. It is well-known that the sequence defined by :
\[
\forall\, k\in\nn,\ y_k=x_k-B,\text{ where } B=(Id-A)^{-1} b
\]
satisfies:
\[
\forall\, k\in\nn,\ y_{k+1}=A y_k. 
\]
Finally, we have :
\[
\forall\, k\in\nn,\ x_k=A^k y_0+B
\]
This latter expression leads to a new formulation of Problem~\eqref{pbopt}: 
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} \left(A^k y_0+B\right)^\intercal Q  \left(A^k y_0+B\right)+q^\intercal  \left(A^k y_0+B\right)
\]
or
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} y_0^\intercal (A^k)^\intercal Q  A^k y_0+(2B+q)^\intercal A^k +B^\intercal Q B+q^\intercal B
\]
We conclude that we can the results developed in Subsection~\ref{mainsub} where the matrix $Q$ is unchanged and the vector $q$ to use is now $2B+q$. The polytope of initial conditions also changes since we have to consider now $\xin-B$. We have to adapt Assumption~\ref{assumR} with  and the new  $2B+q'$ and $\xin-B$.

\section{Practical computations of $K(t,P)$}
\label{computations}
The computations have been performed using Matlab. We use classical internal routines for powers of matrices  (needed for $A^{\ks}$) and  the vector Euclidean norm (needed for $\norm{q}_2$). We discuss here the computations that rely on semi-definite programming. The solver used in the development is Mosek interfaced with Yalmip.

To compute $\bigk$, we need to solve two non-convex problems on the cone of positive semi-definite matrices:
$\inf\{K(1,P)\mid P\in\lyaun\}$ and $\inf\{K(\lmax{P^{-1/2}QP^{-1/2}},P)\mid P\in\lyap{A}\}$. The non-convexity of the two objective functions do not allow use to use classical solvers two compute these two integers. For the moment, to solve exactly the problem is left open. Consequently, in practice, we compute an element of $P_1\in\lyaun$ and an element of $P_0\in\lyap{A}$ and take the minimum between $K(1,P_1)$ and $K(\lmax{P_0^{-1/2}QP_0^{-1/2}},P_0)$. The elements $P_0$ and $P_1$  are computed as optimal solutions of linear semi-definite programs. We tried three objective functions :
\begin{align}
&F_0(P):=\sup_{x\in\xin} x^\intercal P x\\
&F_1(P):=\sup_{x\in\xin} x^\intercal (P-Q) x\\
&F_2(P):=\lmax{P}
\end{align} 
%First we remark that the computation of $\bigk$ needs to solve two minimization problems $\inf\{K(1,P)\mid P\in\lyaun\}$ and 
%$\inf\{K(\lmax{P^{-1/2}QP^{-1/2}},P)\mid P\in\lyap{A}\}$. To  
%
%To compute an interesting integer $K(t,P)$, we tried three objective functions to compute a matrix $P$. Indeed, we do not compute $t$ and $P$ in the same time. Either, we compute $t$ before $P$ or the converse. If $t$ is computed first, we have $t=1$ following the second assertion of Prop.~\ref{comp-prop} and we compute a $P\in\lyaun$. It $P$ is computed first as an element of $ \lyap{A}$, then $t=\lmax{P^{-1/2}QP^{-1/2}}$ following the first assertion of Prop.~\ref{comp-prop} . These two 
%
%In the computations of the two integers, we need first a matrix that satisfies the discrete Lyapunov equation. This matrix equation 
%appears as a constraint in the semi-definite programs. In this paper, we propose three objective convex functions on the convex cone
%of semi-definite matrix  :
%
%For the computation of $K(1,P)$, we need $P \in \lyaun$. Twhere $\varepsilon$ is set to $0.01$.hus, 
%
%We propose to compute this matrix $P$ as an optimal solution  of the semi-definite problem:

For each objective functions, we add the desired constraints (up to perturbation $\varepsilon=0.01$ to ensure the strict positiveness of discrete Lyapunov equation matrix constraint), we get two families of problems parameterized by $i=0,1,2$:

@@

\begin{equation}
\label{eq:klmaxp}
\begin{array}{llc}
\Min & & F_i(P)\\
 & \st &\left\{\begin{array}{l} 
         P-A^\intercal P A-\varepsilon Id\succeq 0\\
         P\succeq  0
        \end{array}\right.
\end{array}
\end{equation}

We will write $P_1$ for an optimal solution of Pb.~\eqref{eq:kunp} and $\pmax$ for an optimal solution of Pb.~\eqref{eq:klmaxp}.

@@

Now, for both $P_1$ and $\pmax$ we have to compute $\lmin{\cdot}$, $\mu(\cdot)$ and $\norm{A}_{\cdot}$. Folowing Lemma~\ref{lemma1}, since $P_1$ and $\pmax$ are symmetric, the minimal eigenvalue
of those matrices can be computed from semi-definite programming. Indeed for all symmetric matrix $M$, we have : 
\[
\lmin{M}=\inf\{\alpha>0\mid M-\alpha Id\geq 0\}
\] 
To compute $\mu(P_1)$ and $\mu(\pmax)$ we suppose that number of vertices of $\xin$ is small and we evaluate directly $\mu(\cdot)$ by enumeration of the image of vertices of $\xin$ and choose the maximal value in the finite list. In a future work, we consider scalable techniques to compute the exact value of $\mu(\cdot)$. The square root (as it is increasing) involved $\mu(\cdot)$ can be performed at the end of the maximization process: we compute the values $x^\intercal P_1 x$ (resp. $x^\intercal \pmax x$) for all vertices; take the maximum of all of them and finally take the square root of the maximal value.

@@

To complete the computations, we have to deal with $\lfrak$. We suppose that the integer $\ks$ is given and then to compute $\lfrak$ we employ the same technique used for $\mu(\cdot)$. It suffices to browse the vertices of $\xin$ once since we evaluate, for each vertice, $\min\{x^\intercal Q x, x^\intercal {A^{\ks}}^\intercal Q {A^{\ks}}x+q^\intercal A^{\ks} x\}$ and take the maximal value. 
\section{Experiments}
\label{experiments}
We decompose the experiments into two sub-classes: the linear and affine systems. Those sub-classes are separeted in three parts : the  case where the objective function is a quadratic homogeneous function; the linear case and the non-homogeneous quadratic objective function. For each cut, we provide a numerical example.  
\subsection{Linear systems}

\subsubsection{Homogeneous objective function}
\label{exp-linear}
We consider the discretisation of an harmonic oscillator $\ddot{x}+\dot{x}+x=0$ by an explicit Euler scheme. The discretization step $h$ is set to 0.01. Introducing the position variable, $x$ and the speed variable $v$. We assume that the initial conditions can be taken into the set $[-1,1]^2$. The Euler scheme becomes a linear discrete-time system in dimension two defined as follows: 
\[
\begin{pmatrix}
x_{k+1}\\
v_{k+1}
\end{pmatrix}
=\begin{pmatrix}
1 & h\\
-h & 1-h
\end{pmatrix}
\begin{pmatrix}
x_{k}\\
v_{k}
\end{pmatrix},\ (x_0,v_0)\in [-1,1]^2
\]
We want to prove that for all $k\in\nn$, $x_{k}^2$ and $v_{k}^2$ are less than 1.

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Numbers/Objective functions & $F_P$ & $F_{P,Q}$ & $F_{\rm max}$ \\
\hline
$K_1$ & & &  \\
\hline
 $K_{\rm max}$ & & &  \\
\hline
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Numbers/Objective functions & $F_P$ & $F_{P,Q}$ & $F_{\rm max}$ \\
\hline
$K_1$ & & &  \\
\hline
 $K_{\rm max}$ & & &  \\
\hline
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Numbers/Objective functions & $F_P$ & $F_{P,Q}$ & $F_{\rm max}$ \\
\hline
$K_1$ & & &  \\
\hline
 $K_{\rm max}$ & & &  \\
\hline
\end{tabular}
\end{center}

\subsubsection{Non-Homogeneous objective function}
We consider a property set of the form $C=\{x\in\rd\mid c^\intercal x\leq \beta,\ c^\intercal x\leq \alpha \}$ with $\alpha\leq \beta$. The latter $C$ can be rewritten as $\{x\in\rd\mid (c^\intercal x)^2-(\beta+\alpha) c^\intercal x\leq -\alpha\beta\}$. Assumption~\eqref{assumR} is true if and only $\alpha\beta< 0$ and thus only valid if $\alpha<0$. A MIEUX EXPLIQUER

Let us consider the same linear system as in Subsubsection~\ref{exp-linear}. We are interested in proving that for all $k\in\nn$, $-1\leq x_k-0.5v_k\leq 2$. This is the same as proving 
$(x_k-0.5 v_k)^2-x_k+0.5 v_k\leq 1$.
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Numbers/Objective functions & $F_P$ & $F_{P,Q}$ & $F_{\rm max}$ \\
\hline
$K_1$ & & &  \\
\hline
 $K_{\rm max}$ & & &  \\
\hline
\end{tabular}
\end{center}
\subsection{Affine systems}

>>>>

We can represent the evolution of an affine system by the following relation:
\begin{equation}
\label{context}
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k + b  
\end{equation}
where:
\begin{itemize}
\item $A$ is a non-zero square matrix of size $d\times d$;
\item $b$ is vector of $\rd$;
\item  $\xin$ is a non-empty polytope (bounded polyhedra) non reduced to one vector. 
\end{itemize}
We insist on the fact that the initial values are represented by an infinite bounded set. It can be interpreted as a set of perturbations, non-determinism, different cases...

@@

For the matrix $A$ of the affine system considered at Eq.~\eqref{context}, we introduce the set $\lyap{A}$ of the  solutions  of the discrete Lyapunov equation :
\[
\lyap{A}=\{P\in\sym \mid P\succ 0,\ P-A^\intercal P A\succ 0\}\enspace .
\]
Hence, $\rho(A)<1$ ensures that $\lyap{A}$ is non-empty. More precisely, $\rho(A)<1$ if and only if for all positive definite matrices $R$ there exists a unique positive definite matrix $P$ such that $P-A^\intercal P A=R$.

@@

We recall Weyl's inequalities~\cite{horn1990matrix} that provide inequalities for the eigenvalues of the sum of two symmetric matrices and the sum of the eigenvalues of each matrix.  
\begin{lemma}[Weyl's inequalities]
\label{lemmaWeyl}
Let $M$ and $N$ be symmetric matrices. We have, for all $\ell\in\{1,\ldots,d\}$ :
\[
\lambda_\ell(M)+\lmin{N}\leq \lambda_\ell(M+N)\leq \lambda_\ell(M)+\lmax{N}
\]
\end{lemma}
The following lemma recalls that, when we maximize a convex function over a polytope, it suffices to consider the values of the function at the extreme points. 
\begin{lemma}[Maximisation of a convex function over a polytope]
\label{lemma2}
Let $C$ be a polytope and $f:\rd\to\rd$ a convex function. Then:
\[
\max_{x\in C} f(x)=\max_{x\in \mathcal{E}(C)} f(x)
\]
where $\mathcal{E}(C)$ denotes the finite set of the extreme points (vertices) of $C$.
\end{lemma}
Note that the result also holds for convex compacts sets.  
\subsection{Verification Linear Time-Invariant Discrete-time Systems}
\label{mainsub}
Now we come back to the verification problem presented at~\eqref{pbopt}. In this subsection, we suppose that the system in linear i.e. $b=0$ in~\eqref{context}. We will show at Subsection~\ref{affine} that the case $b\neq 0$ can be reduced to the linear case.

@@

We will need the following notation :
\[
\ks:=\inf\left\{k\in\nn\mid \min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\right\}>0\right\}
\]
and
\[
\lfrak:= \min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^{\ks})^\intercal Q A^{\ks} x+q^\intercal A^{\ks}x\right\}
\]

@@

\begin{prop}
\label{interiorprop}
If $\xin$ contains $0$ in its interior then Assumption~\ref{assumR} holds. In this case, $\ks=0$.
\end{prop}

\begin{proof}
Suppose that $q\neq 0$. Since 0 belongs to the interior of $\xin$, then it exists $\gamma>0$ such that $[-\gamma,\gamma]^d\subset \xin$. Let $x=\gamma q (2\norm{q}_{\infty})^{-1}$. Then $q^\intercal x=\norm{q}_2^2\norm{q}_{\infty}^{-1} >0$ and $x\in [-\gamma,\gamma]^d$. From, $Q\succeq 0$, $x^\intercal Q x+q^\intercal x>0$. We thus have $\sup_{y\in \xin}  y^\intercal Q y+q^\intercal y>0$.

Now, since $Q$ is not null, there exists $z\in\rd$ such that $z\neq 0$ and $z^\intercal Q z>0$. Taking $x=\gamma z (2\norm{z}_\infty)^{-1}$, we get $x^\intercal Q x=\gamma ^2 (2\norm{z}_\infty)^{-2} z^\intercal Q z>0$ and $x\in[-\gamma,\gamma]^d$. Then $\sup_{y\in\xin} y^\intercal Q y>0$.

Finally, $\displaystyle{\min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal Q x+q^\intercal x\right\}>0}$ and $\ks=0$.
\end{proof}

Prop.~\ref{interiorprop} uses the fact that there exists some $k\in\nn$ and some $x\in\xin$ for which all coordinates of $A^k x$ have the same sign as the ones of $q$. A more general proposition can be written using this fact. However, the existence of such $k$ and such $x$ cannot be decided.

\begin{remark}
\label{remarklinear}
We can apply our method when the property is defined from linear forms. 
Let us consider the following special case. Let $\alpha,\beta\in\rr$ such that $\alpha<\beta$, and $c\in\rd$. The set $\{x\in\rd \mid \alpha \leq c^\intercal x\leq \beta\}$ can be rewritten as:
$x^\intercal (c c^\intercal) x-(\beta+\alpha) c^\intercal x \leq -\alpha \beta$. Then we have in this case $Q=c c^\intercal$ and $q= -(\beta+\alpha) c$. With these data, if Assumption~\ref{assumR} holds then $\alpha$ is negative and $\beta$ is positive. 
\end{remark}

@@

\begin{prop}
\label{lyapunovbigk}
Suppose that $q=0$ and $Q\in\lyap{A}$ then $K(1,Q)=1$.  
\end{prop}

\begin{proof}
Since $Q\in\lyap{A}$ then $Q\succ 0$ and thus $\displaystyle{\sup_{x\in\xin} x^\intercal Q x>0}$ and $\ks=0$. We have $\lfrak=\displaystyle{\sup_{x\in\xin} x^\intercal Q x}$. Hence, since 
$\sqrt{\cdot}$ is increasing then $\sqrt{\lfrak}=\displaystyle{\sup_{x\in\xin} \sqrt{x^\intercal Q x}}$. Finally, $K(1,Q)=1+\displaystyle{\ln\left(\sqrt{\lfrak}/\sqrt{\lfrak}\right) / \ln(\norm{A}_Q)}=1$.
\end{proof}

@@

The result presented in Lemma~\ref{lyapnorm} explains the choice of the matrix norm. It is well-known that $\rho(A)<1$ is equivalent to the existence to a matrix norm $\norm{\cdot}$ for which $\norm{A}<1$. Here we exhibit a compatible norm. We could choose another norm but we would have to restrict the class of systems that we could analyze.

@@

%&=&\displaystyle{\max\left\{\max_{0\leq k\leq \bigk-1}\max_{x\in \mathcal{E}(\xin)} \nu_k(x),\lfrak\right\}}
%\end{array}
%\]
Now, for all $k< \ks$, for all $x\in\xin$, $\nu_k(x)\leq 0$ and for all $x\in\xin$, $\nu_{\ks}(x)\leq \lfrak$. Moreover, from Th~\ref{thfond}, 
for all $k>k^*$, $\nu_k(x)\leq \lfrak$. The conclusion follows from $\displaystyle{\max_{x\in\xin} \max_{\ks\leq k\leq k^*} \nu_k(x)\geq \max_{x\in\xin} \nu_{\ks}(x)\geq \lfrak}$. 
%The conclusion follows from the fact that  $\lfrak\leq \max_{x\in \mathcal{E}(\xin)} \nu_{\ks}(x)$.
\end{proof}
The integer $\bigk$ is the best integer than we can get in theory. In practice, we cannot compute $\bigk$. We will solve others semi-definite problems but we will still use the function $(t,P)\mapsto K(t,P)$. We will describe the practical approach at Section~\ref{computations}.

%since we cannot solve the minimization problems :
%\[
%\inf_{P\in\lyaun} K(1,P)\text{ and }\inf_{P\in\lyap{A}}K(\lmax{P^{-1/2} Q P^{-1/2}},P)
%\]
%%In practice, for the moment, we only compute two integers : $K( 1,P)$ where $P$ is an optimal solution of a perturbation of the minimization problem : 
%%\begin{equation}
%%\label{minlmaxQ}
%%\inf\{\lmax{P}\mid P\in \lyaun\}
%%\end{equation}
%%and $K(\lmax{P^{-1/2} Q P^{-1/2}},P)$ where $P$ is an optimal solution of a perturbation of the minimization problem :
%%\begin{equation}
%%\label{minlyap}
%%\inf\{\lmax{P}\mid P\in\lyap{A}\}\enspace .
%%\end{equation}
%Then we take the minimum of the  two integers. We come back latter on the perturbation of the two minimization problems at Section~\ref{computations}.
\subsection{From linear systems to affine ones}
\label{affine}
We come back to Problem~\eqref{context}: the case where the system is affine ($b\neq 0$).
A possible natural reformulation of affine systems into linear one is to use lift-and-project. However, lift-and-project would add 1 as eigenvector of the matrix defining the new linear system. Assumption~\ref{assum2} would be violated.  Consequently, we adopt another basic approach which consists in using an auxiliary linear discrete-time system.

 We recall that we have :
%Let us consider the same problem where we replace the linear dynamics by an affine one:
\[
x_0\in \xin,\ \forall\, k\in\nn,\ x_{k+1}=A x_k+b,\
\]
where $A$ is a $d\times d$ matrix and $b$ a vector of $\rd$. We recall that $\rea$ is given by Formula~\eqref{reach} and we consider the optimization problem:
\[
\sup_{x\in\rea} x^\intercal Q x+q^\intercal x
\]
Assumption~\ref{assum2} still holds and it implies that $Id-A$ is invertible. It is well-known that the sequence defined by :
\[
\forall\, k\in\nn,\ y_k=x_k-B,\text{ where } B=(Id-A)^{-1} b
\]
satisfies:
\[
\forall\, k\in\nn,\ y_{k+1}=A y_k. 
\]
Finally, we have :
\[
\forall\, k\in\nn,\ x_k=A^k y_0+B
\]
This latter expression leads to a new formulation of Problem~\eqref{pbopt}: 
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} \left(A^k y_0+B\right)^\intercal Q  \left(A^k y_0+B\right)+q^\intercal  \left(A^k y_0+B\right)
\]
or
\[
\sup_{k\in\nn} \sup_{y_0\in\xin-B} y_0^\intercal (A^k)^\intercal Q  A^k y_0+(2B+q)^\intercal A^k +B^\intercal Q B+q^\intercal B
\]
We conclude that we can use the results developed in Subsection~\ref{mainsub} where the matrix $Q$ is unchanged and the vector $q$ to use is now $2B+q$. The polytope of initial conditions also changes since we have to consider now $\xin-B$. We have to adapt Assumption~\ref{assumR} with the new linear part equal to $2B+q$ and the new initial polytope equal to $\xin-B$.

\section{Practical computations of $K(t,P)$}
\label{computations}
The computations have been performed using Matlab. We use classical internal routines for powers of matrices (needed for $A^{\ks}$), the vector Euclidean norm (needed for $\norm{q}_2$) and the inverse of Id-A. We discuss here the computations that rely on semi-definite programming. The solver used in the development is Mosek~\cite{andersen2000mosek} interfaced with Yalmip~\cite{Lofberg2004}.

To compute $\bigk$, we need to solve two non-convex problems on the cone of positive semi-definite matrices:
\[
\inf\{K(1,P)\mid P\in\lyaun\} \text{ and }\inf\{K(\lmax{P^{-1/2}QP^{-1/2}},P)\mid P\in\lyap{A}\} \enspace.
\]

The non-convexity of the two objective functions do not allow us to use classical solvers to compute these two integers. For the moment, to solve exactly the two problems is left open. Consequently, in practice, we compute an element of $P_1\in\lyaun$ and an element of $P_0\in\lyap{A}$ and take the minimum between $K(1,P_1)$ and $K(\lmax{P_0^{-1/2}QP_0^{-1/2}},P_0)$. The elements $P_0$ and $P_1$  are computed as optimal solutions of linear semi-definite programs. We tried five objective functions :
\begin{align}
&F_0(P):=\sup_{x\in\xin} x^\intercal P x\\
&F_1(P):=\sup_{x\in\xin} x^\intercal (P-Q) x\\
&F_2(P):=\sum_{x\in\xin} x^\intercal P x\\
&F_3(P):=\sum_{x\in\xin} x^\intercal (P-Q) x\\
&F_4(P):=\lmax{P}
\end{align} 
%First we remark that the computation of $\bigk$ needs to solve two minimization problems $\inf\{K(1,P)\mid P\in\lyaun\}$ and 
%$\inf\{K(\lmax{P^{-1/2}QP^{-1/2}},P)\mid P\in\lyap{A}\}$. To  
%
%To compute an interesting integer $K(t,P)$, we tried three objective functions to compute a matrix $P$. Indeed, we do not compute $t$ and $P$ in the same time. Either, we compute $t$ before $P$ or the converse. If $t$ is computed first, we have $t=1$ following the second assertion of Prop.~\ref{comp-prop} and we compute a $P\in\lyaun$. It $P$ is computed first as an element of $ \lyap{A}$, then $t=\lmax{P^{-1/2}QP^{-1/2}}$ following the first assertion of Prop.~\ref{comp-prop} . These two 
%
%In the computations of the two integers, we need first a matrix that satisfies the discrete Lyapunov equation. This matrix equation 
%appears as a constraint in the semi-definite programs. In this paper, we propose three objective convex functions on the convex cone
%of semi-definite matrix  :
%
%For the computation of $K(1,P)$, we need $P \in \lyaun$. Twhere $\varepsilon$ is set to $0.01$.hus, 
%
%We propose to compute this matrix $P$ as an optimal solution  of the semi-definite problem:

For each objective functions, we add the desired constraints (up to perturbation $\varepsilon=0.01$ to ensure the strict positiveness of discrete Lyapunov equation matrix constraint), we get two families of problems parameterized by $i=0,1,2,3,4$:

@@

\begin{equation}
\label{eq:klmaxp}
\begin{array}{llc}
\Min & & F_i(P)\\
 & \st &\left\{\begin{array}{l} 
         P-A^\intercal P A-\varepsilon Id\succeq 0\\
         P\succeq  0
        \end{array}\right.
\end{array}
\end{equation}
The problems ~\eqref{eq:kunp} and~\eqref{eq:klmaxp} with the objective functions $F_0$ and $F_1$ can be rewritten as linear semi-definite programs. Indeed they are equivalent to minimize $t$ such that $t\geq x^\intercal P x$ for all $x\in\mathcal{E}( \xin)$ ($P$ being constrained to be semi-definite positive). Note also that the constraint $P-A^\intercal P A-\varepsilon Id\succeq 0$ implies that $P$ is actually positive definite.  
We will write $P_1$ for an optimal solution of Pb.~\eqref{eq:kunp} and $\pmax$ for an optimal solution of Pb.~\eqref{eq:klmaxp}.

@@

Now, for both $P_1$ and $\pmax$ we have to compute $\lmin{\cdot}$, $\mu(\cdot)$ and $\norm{A}_{\cdot}$. Following Lemma~\ref{lemma1}, since $P_1$ and $\pmax$ are symmetric, the minimal eigenvalue
of those matrices can be computed from semi-definite programming. Indeed for all symmetric matrix $M$, we have : 
\[
\lmin{M}=\inf\{\alpha>0\mid M-\alpha Id\geq 0\}
\] 
To compute $\mu(P_1)$ and $\mu(\pmax)$ we suppose that the number of vertices of $\xin$ is small and, following Lemma~\ref{lemma2}, we evaluate directly $\mu(\cdot)$ by enumeration of the image of vertices of $\xin$ and choose the maximal value in the finite list. In a future work, we consider scalable techniques to compute the exact value of $\mu(\cdot)$. The square root (as it is increasing) involved $\mu(\cdot)$ can be performed at the end of the maximization process: we compute the values $x^\intercal P_1 x$ (resp. $x^\intercal \pmax x$) for all vertices; take the maximum of all of them and finally take the square root of the maximal value.

@@

To complete the computations, we have to deal with $\lfrak$. We suppose that the integer $\ks$ is given and then to compute $\lfrak$ we employ the same technique used for $\mu(\cdot)$. It suffices to browse the vertices of $\xin$ once since we evaluate, for each vertex, $\min\{x^\intercal Q x, x^\intercal {A^{\ks}}^\intercal Q {A^{\ks}}x+q^\intercal A^{\ks} x\}$ and take the maximal value. 
\section{Experiments}
\label{experiments}
We decompose the experiments into two sub-classes: the linear and affine systems. Those sub-classes are separeted in three parts : the  case where the objective function is a quadratic homogeneous function; the linear case and the non-homogeneous quadratic objective function. For each cut, we provide a numerical example.  
\subsection{Linear systems}

\subsubsection{Homogeneous objective function}
\label{exp-linear}
We consider the discretisation of an harmonic oscillator $\ddot{x}+\dot{x}+x=0$ by an explicit Euler scheme. The discretization step $h$ is set to 0.01. Introducing the position variable, $x$ and the speed variable $v$. We assume that the initial conditions can be taken into the set $[-1,1]^2$. The Euler scheme becomes a linear discrete-time system in dimension two defined as follows: 
\[
\begin{pmatrix}
x_{k+1}\\
v_{k+1}
\end{pmatrix}
=\begin{pmatrix}
1 & h\\
-h & 1-h
\end{pmatrix}
\begin{pmatrix}
x_{k}\\
v_{k}
\end{pmatrix},\ (x_0,v_0)\in [-1,1]^2
\]
We want to prove that :
the trajectories are bounded and compute a bound over the Euclidean norm of the state-variable; for all $k\in\nn$, $x_{k}^2$ and $v_{k}^2$ are less than 1.

\begin{enumerate}
\item Boundedness. The boundedness is already proved since $\rho{A}<1$ and $A^k x$ tends to 0 as $k$ tends to $+\infty$. However, we do not have a bound over the value taken by the state-variable.  
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$\norm{(x_k,v_k)}^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

\item The proof of $x_{k}^2\leq 1$ :

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$x_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}

\item The proof of $v_{k}^2\leq 1$

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\multicolumn{6}{|c|}{$v_k^2$}\\
\hline
Numbers/Objective functions & $F_0$ & $F_{1}$ & $F_{2}$ & $F_3$ & $F_4$ \\
\hline
$K_1$ & & & & & \\
\hline
 $K_{\rm max}$ & & & & & \\
\hline
\end{tabular}
\end{center}
\end{enumerate}
